FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-15 paper.tex
Use Julia for examples of floating-point inaccuracy

% If using Emacs, put cursor on last closing paren and call "eval-last-sexp" (usually `C-x C-e') to regenerate
% (save-excursion
%  (let* ((toc (funcall imenu-create-index-function))
%         (headers (mapcar (pcase-lambda (`(,str . ,loc)) str)
%                          toc)))
%    (beginning-of-buffer)
%    (search-forward (concat "%% " "--- autogenerated TOC ---"))
%    (next-line)
%    (beginning-of-line)
%    (let ((start (point)))
%      (search-forward (concat "%% " "--- end autogenerated TOC ---"))
%      (beginning-of-line)
%      (delete-region start (point))
%      (goto-char start))
%    (cl-loop for header in headers
%             do (insert (format "%% %s\n" header)))
%    (next-line)))

%% --- autogenerated TOC ---
% juliacon
% document
%   Introduction
%   Floating Point Exception Primer
%     Where do \NaN
%       TODO good quote, try to work this in
%     Infinity
%     To Kill a Floating Point
%   \FlowFPX
%     \FT
%       Tracking Excptional Values
%       Fuzzing
%     Visualization
%     \GPUFPX
%   Case Studies
%     Surprise \NaN
%       CSTGs help us get a bigger picture
%     \NaN
%     Finch
%     A Bug-Hunting Safari
%     \GPUFPX
%   Related Work
%   Discussion
%     Tracking more than \NaN
%     Enhanced Fuzzing
%     Static FPX, Gradual Types
%   Acknowledgments
% document
%% --- end autogenerated TOC ---

@@

\Fp{} is essentially scientific notation---\fp{} numbers are represented by packing a sign bit, an exponent, and an fraction part (also called the ``significand'' or ``mantissa'') into contiguous bit fields.
The advantages of this representation are that allows for compact representation of both very small and very large numbers in a narrow range of bits---a 64-bit float can represent numbers as big as $2^{1024}$ and as small as $2^{-1074}$.

This range is sufficient to represent both Planck's constant as well as Avogadro's constant with reasonable accuracy, making IEEE 754 \fp{} adequate for a wide range of numerical computations.
The flip side is that not every real number within that wide range is representable as a \fp{} value---there is almost always some gap between the real number that we wish to represent and the \fp{} number that can represent.
Whenever a value we have doesn't have a precise representation, we must \emph{round} that real value to the nearest \fp{} value.
We call this gap error, and error, along with the finite nature of \fp{} values, means that we can run into some funny situations.

For example, if two numbers' exponents differ by too large an amount, adding or subtracting the smaller from the bigger can result in just the bigger number.
For example, $6.022 \times 10^{23} + 6.626 \times 10^{-34}$ (Avogadro's constant plus Planck's constant) will leave with with just Avogadro's constant---there is no way to pack all the significant figures from true value that differs minutely from Avogadro's number into a 64-bit \fp{} number.
This is a fairly extreme example; even when the numbers are closer in magnitude, most operations on \fp{} numbers induce loss of accuracy.

>>>>

@@

\Fp{} is essentially scientific notation---\fp{} numbers are represented by packing a sign bit, an exponent, and an fraction part (also called the ``significand'' or ``mantissa'') into contiguous bit fields.
The advantages of this representation are that allows for compact representation of both very small and very large numbers in a narrow range of bits---a 64-bit float can represent numbers as big as $2^{1024}$ and as small as $2^{-1074}$.
This range is sufficient to represent numbers as small as Planck's constant or as big as Avogadro's constant with reasonable accuracy.

The flip side is that not every real number within that wide range is representable as a \fp{} value---there is almost always some gap between the real number that we wish to represent and the \fp{} number that can represent.
Whenever a value we have doesn't have a precise representation, we must \emph{round} that real value to the nearest \fp{} value.
We call this gap error, and error, along with the finite nature of \fp{} values, means that we can run into some funny situations.

For example, if two numbers' exponents differ by too large an amount, adding or subtracting the smaller from the bigger can result in just the bigger number.
Below we take the Planck constant and add it to Avogadro's constant and are left bit-for-bit with Avogadro's constant, as if we had never done any addition at all: there is no way to pack all the significant figures from true value that differs minutely from Avogadro's number into a 64-bit \fp{} number.

\begin{lstlisting}[language = Julia]
p = 6.62607015e-34
a = 6.02214076e23

println("planck:   $p : $(bitstring(p))")
println("avogadro: $a  : $(bitstring(a))")
println("p + a:    $(p + a)  : $(bitstring(p + a))")
\end{lstlisting}

This outputs:

\begin{lstlisting}
planck:   6.62607015e-34 : 0011100100001 ... 001
avogadro: 6.02214076e23  : 0100010011011 ... 111
p + a:    6.02214076e23  : 0100010011011 ... 111
\end{lstlisting}

This is a fairly extreme example; even when the numbers are closer in magnitude, most operations on \fp{} numbers induce loss of accuracy.
